Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-08 paper/ijm.tex
ijm: written till conclusion

\def\tenpc{$^{\ast}$}
\def\fivepc{$^{\ast\ast}$}
\def\onepc{$^{\ast\ast\ast}$}
\newcommand{\legend}{\normalsize{Significance levels:\hspace{1em} \tenpc : 10\% \hspace{1em} \fivepc : 5\% \hspace{1em} \onepc : 1\% \normalsize}}
\def\sep{0.5em}
\def\legendTwo{\multicolumn{3}{l}{\footnotesize{Significance levels
			:\hspace{1em} $\dag$ : 10\% \hspace{1em}
			$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%SKB: Added for documenting the tables and figures
\newcommand{\modelTwo}{age, age\textsuperscript{2}, education, marriage, number of children, inter-ethnic household}
\newcommand{\modelThree}{\modelTwo, industry, occupation, ownership type of company, number of workers in company, working in the public sector, experience in company}
\newcommand{\modelThreeAdd}{industry, occupation, ownership type of company, number of workers in company, working in the public sector, experience in company}	
\newcommand{\restrictions}{The sample period is from year 2000 to year 2012. Sample is limited to persons
	25-55 year old.}
\newcommand{\agerestrictions}{25-55 year olds only.}
%Modified section numbering	
\renewcommand{\thesection}{\arabic{section}.}
\renewcommand{\thesubsection}{\thesection\arabic{subsection}.}

@@

We also estimate the effects separately by age groups and time in
order to detect potential trends across the age groups and across ethnic groups
(Table~\ref{tab:age_group_trend}).
We look separately at 4 sub-periods: Early transition period
(1989--1995); First economic growth period (1996--1999);
Economic boom, EU accession (2000--2007); and the recession
(2008--2012).

We use the same econometric specification as described in section \ref{sec:method}, the results presented here are derived using the model 2.

For ethnic Estonians, knowledge of Russian
(Table~\ref{tab:age_group_trend}, upper panel) has become less
important after its peak in the late 1990s. Notably, the
association between wage and language skills are small in the
young groups, the age groups where the Russian skills are also rapidly declining.
Russian is still closely related to salary for
older workers. The importance of English language skills fluctuates
and declines slowly through the period, just the oldest male group
(45-55) and the youngest female group (25-35) show a growing trend.
This potentially reflects both the changing supply and demand of
English skills.

Now we turn to ethnic Russians (lower panel in the table).  The association between Estonian skills and income shows no
statistically significant results at all.  We'll return to this
outcome below.
The positive impact of Estonian language skills for Russian women
increases
over the sample period for the young group (25--35)
and the prime age group (36--45).
Good
Estonian language knowledge is associated with approximately 10\% higher salary for Russian women for the boom in the 2000s. For the
recession period, the figure is even close to 12\%. These age
groups are the workers who were in their twenties or younger during
the major political and economic transition in early 1990s. It
suggests that the previously segregated ``Estonian'' and ``Russian''
economy are slowly becoming a more integrated one where ethnic
Russians, at least women, with a good Estonian command can get well-paid jobs.
In comparison, the estimates for English language fluency shows
downward trend for all age groups for men and for women.  A likely
cause is the increasing abundance of these skills over time.

@@

We repeat the exercise for unemployment models (using specification 2
in
Table~\ref{tab:unemployment_estimation_by_sex_and_ethnic_same_sample}).
However, as most of the results are not statistically significant, we
do not present the tables here.

@@

The most interesting result in our analysis so far--the fact that
Russian men earn virtuannly no Estonian language premium--warrants a
deeper look at the data.  We split the Russian male wage-earners' data according to three
dimensions: Estonian language skills, residence in Tallinn metro area,
and education.  However, to correct very strong trends through these
years both in wage and language skills, we present the average
residuals from an OLS regression that explains log wage by year fixed
effects (Table~\ref{tab:loc-lang-edu}). 
\fxwarning{ot: Sven, can you redo this table?  Use exactly the same
  sample as you used.  Would also be good to put both et1 and et0 on
  the same row}
The table show that the picture is in fact more complex.  We can see
that there is little to no gain from Estonian skills for the
largest group, high-school educated workers.  Even more, outside of
the metropolitan area, the the lowest and highest education groups
have in fact negative language premium.  However, these two groups
enjoy quite a noticeable premium in the Tallinn region.  The
opposite-signed effects in for these two education groups suggests
selective internal migration may play a role.

@@

We can conclude the table~\ref{tab:loc-lang-edu} that our
most controversial finding, no positive effect of Estonian language
skills for Russian men, is primarily driven by workers with
high-school degree.

All previous results can be explained by two hypotheses: first, Russian--speaking men are working
primarily in Russian language environment (ethnic segregation) while this is not true for
women; and second, Russian-speaking men are working in jobs with
little communication requirements (gender segregation).

To shed some additional light on these hypotheses, we use data
from 2008 Integration Monitor Survey.  We focus on
two questions: ``which is the language of communication at your
workplace?'' and ``which languages do you need at work''.  The first
question addresses the ethnic segregation while the second focuses on
the need of communication.  As very few individuals work in English
environment, we only look for Estonian as the communication language.

@@

Table~\ref{tab:environment_descriptive} shows the proportion of
Russian respondents who use Estonian as the main communication language at work, and
who need Estonian and English at work.  As we see, women are more likely to
use Estonian than men while the opposite is true for English.
Table~\ref{tab:environment_model} presents the regression results when
controlling for other
characteristics, including education and geographic location.

@@

The table indicates that Russian women need Estonian 26 percentage
points more at work than Russian men.  The table also suggests that
Russian women are 16 percentage points more likely to be working in
Estonian-speaking environment when controlling for other factors.
Both estimates are statistically significant (at 5\% and 10 \% level
respectively), but not precisely estimated.  We can also see that
college education is associated by much more Estonian need and usage,
but in regions where Russian speakers are concentrated, the capital
Tallinn and the North-East, Estonian is less needed.  In contrast, no
gender difference is visible for English needs.

In summary, this dataset supports the previous hypothesis: Russian women are
working in less segregated environment where Estonian is the main
communication language, and they need Estonian for other reasons as
well.  Gender differences in English usage are not statistically
significant.
\fxwarning{ot: let's remove the cv.ee stuff.  Or substantially improve
that.}

>>>>

\def\tenpc{$^{\ast}$}
\def\fivepc{$^{\ast\ast}$}
\def\onepc{$^{\ast\ast\ast}$}
\newcommand{\legend}{\normalsize{Significance levels:\hspace{1em} \tenpc : 10\% \hspace{1em} \fivepc : 5\% \hspace{1em} \onepc : 1\% \normalsize}}
\def\sep{0.5em}
\def\legendTwo{\multicolumn{3}{l}{\footnotesize{Significance levels
			:\hspace{1em} $\dag$ : 10\% \hspace{1em}
			$\ast$ : 5\% \hspace{1em} $\ast\ast$ : 1\% \normalsize}}}
%SKB: Added for documenting the tables and figures
\newcommand{\modelTwo}{age, age\textsuperscript{2}, education, marriage, number of children, inter-ethnic household}
\newcommand{\modelThree}{\modelTwo, industry, occupation, ownership type of company, number of workers in company, working in the public sector, experience in company}
\newcommand{\modelThreeAdd}{industry, occupation, ownership type of company, number of workers in company, working in the public sector, experience in company}	
\newcommand{\restrictions}{The sample period is from year 2000 to year 2012. Sample is limited to persons
	25-55 year old.}
\newcommand{\agerestrictions}{25-55 year olds only.}
% Modified section numbering: removed dot at the end as this jumps
% into the cross-references.
\renewcommand{\thesection}{\arabic{section}}
\renewcommand{\thesubsection}{\thesection\arabic{subsection}}

@@

As age is a central determinant how major economic and political
reforms influence individual fates, 
we also analyze the time trends by age groups
(Table~\ref{tab:age_group_trend}).
We group the observation period into 4 sub-periods: Early transition period
(1989--1995); First economic growth period (1996--1999);
Economic boom, EU accession (2000--2007); and the recession
(2008--2012).
We use the same econometric framework as described in section
\ref{sec:method}, the results presented here are based on the specification 2.

For ethnic Estonians, knowledge of Russian
(Table~\ref{tab:age_group_trend}, upper panel) has become less
important after its peak in the late 1990s. Notably, the
association between wage and language skills are small in the
young groups, the age groups where knowledge of that language
declining rapidly.
For older age groups, the estimates are larger but not statistically
significant since the recession period.  English was not important
during the early years of transition, but has been very closely
associated with better pay since 1996.  It's importance seems to be
falling though for the young and middle male groups while the oldest men
(45-55) and the youngest women (25-35) show a growing trend.
This potentially reflects both the changing supply and demand of
English skills.

Now we turn to ethnic Russians (lower panel in the table).  The association between Estonian skills and income shows no
statistically significant results at all.  This is the most unexpected
result in our analysis and we'll return to this
below in section~\ref{sec:segregation_analysis}.
The impact of Estonian language skills for Russian women too, but
since year 2000, we observe a sizable effect of approximately
10\% for several cases.  However, standard errors are large and we
cannot conclude that the effect is positive for all the age groups.
The positive outcomes for women for the second decade in our analysis
suggests that the previously segregated ``Estonian'' and ``Russian''
economies are slowly becoming a more integrated one where ethnic
Russians, at least women, with good Estonian command get better-paid
jobs than those who cannot speak the language.
In comparison, the estimates for English language show
a downward trend for all age groups for both men and for women.  A likely
cause is the increasing abundance of these skills over time.

@@

We repeat the exercise for unemployment models.
However, as most of the results are not statistically significant, we
do not report the results here.

@@

The most surprising result in our analysis so far--the fact that
Russian men earn virtually no Estonian language premium while women do--warrants a
deeper look at the data.  We compute the average wage for the Russian male wage-earners
along the three
dimensions: Estonian language skills, residence in Tallinn metro area,
and education.  The metropolitan area is one of the two major regions
where the Russian-speaking population is concentrated, and it's
booming economy may be very different from that in the North-East.
Even more, while Russians form a large minority in and around Tallinn,
they are in majority in the North-East.  
In addition, to correct very strong wage and language skill trends through these
years, we also present the average
residuals from an OLS regression that explains log wage by year fixed
effects (Table~\ref{tab:loc-lang-edu}). 
\fxwarning{ot: Sven, can you redo this table?  Use exactly the same
  sample as you used.  Would also be good to put both et1 and et0 on
  the same row}
The table shows that the picture is in fact quite complex.  We can see
that there is little to no gain from Estonian skills for the
largest group, high-school educated workers.  Even more, outside of
the metropolitan area, the lowest and highest education groups
have in fact negative language premium.  However, these two groups
enjoy quite a noticeable premium in the Tallinn region.  The
opposite-signed effects in for these two education groups suggests
that a
selective internal migration may play a role.

@@

We can conclude the table~\ref{tab:loc-lang-edu} that our
most suprising finding, no positive effect of Estonian language
skills for Russian men, is primarily driven by workers with
high-school degree.  It also suggests that highly educated workers in
the metropolitan area enjoy, in fact, a substantial language premium.
However, this group is not very large.

The previous results can be explained by two hypotheses: first,
Russian--speaking men, at least those with high-school degree as their
highest completed education, are working
primarily in Russian language environment (ethnic segregation) while this is not true for
women; and second, Russian-speaking men are working in jobs with
little communication requirements (gender segregation).

To shed some additional light on these hypotheses, we use data
from 2008 Integration Monitor Survey.  We focus on
two questions: ``which is the language of communication at your
workplace?'' and ``which languages do you need at work''.  The first
question addresses the ethnic segregation while the second focuses on
the need of communication.

@@

Table~\ref{tab:environment_descriptive} shows the proportion of
Russian respondents who use Estonian as the main communication language at work, and
who need Estonian and English at work.  As we see, women are more likely to
use Estonian while the opposite is true for English.
Table~\ref{tab:environment_model} presents the regression results
where we estimate the likelihood of having the need and using Estonian at work
while
controlling for gender and other individual
characteristics, including education and geographic location.

@@

The table indicates that Russian women need Estonian 26 percentage
points more at work than Russian men.  The table also suggests that
Russian women are 16 percentage points more likely to be working in
Estonian-speaking environment when controlling for other factors.
Both figures are statistically significant (at 5\% and 10 \% level
respectively), but not precisely estimated.  We can also see that
college education is associated by much more Estonian need and usage,
but in regions where Russian speakers are concentrated, the capital
Tallinn and the North-East, Estonian is less needed.  In contrast, no
gender difference is visible for English needs.

In summary, this dataset supports the previous hypothesis: Russian women are
working in less segregated environment where Estonian is the main
communication language, and they need Estonian at work for other reasons as
well.  Gender differences in English usage are not statistically
significant.
\fxwarning{ot: let's remove the cv.ee stuff.  Or substantially improve
that.}
